Disease Knowledge Transfer across Neurodegenerative Diseases

### 2019-01-11 paper/paper.tex
submitted paper to arxiv

\setlength{\textfloatsep}{20pt}
%\setlength{\floatsep}{1pt}
%\setlength{\topmargin}{1pt}

@@

\author{*******************************************}

\institute{***************************************}

@@

\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
\newcommand{\jmdFld}{..}

\begin{abstract}
We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases. DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases. DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases. As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks. Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects. For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD dataset (tAD) from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own centre, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available. We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature. We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches. While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.

@@

We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes. This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology. DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects at risk of diseases. We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data. This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own centre. We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings. We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches. Code for DKT is available online: (URL not shown yet for anonymity).

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
 \caption{Diagram of the proposed DKT framework. We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions. Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region. For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe. The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled. Disease knowledge transfer can then be achieved via the disease-agnostic units. Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
 \label{fig:diagram}
\end{figure}

@@

\begin{equation}
\label{eqDysfunctionScoreDef}
 \gamma_i^l = f(\beta_i + m_{ij}; \lambda_{d_i}^l)
\end{equation}
where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of all agnostic units. Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled. For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.  We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers. Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.

@@

We first test DKT on synthetic data, to assess its performance when ground truth is known. More precisely, we generate data that follows the DKT model exactly, and test DKT's ability to recover biomarker trajectories and subject time-shifts. We generate synthetic data from two diseases (i.e. \emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort. The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$. To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects. Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g. MRI volume) but to represent measurements from different brain regions (e.g. temporal and occipital).

 \begin{table}
 \begin{tabular}{c | c}
& \textbf{Trajectory parameters} \\
%\hline
 Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\

"Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
 "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
\hline
& \textbf{Subject parameters} \\
 Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
 Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
 Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
 Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
\end{tabular}
\caption{Parameters used for synthetic data generation, emulating the TADPOLE dataset and our own dataset.}
\label{tab:synParams}
\end{table}

@@

We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table. From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan. Most subjects also had MRI scans and cognitive tests. In order to model another disease, we further included 76 PCA subjects from our own centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.

For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe. For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own dataset). Finally, we normalised the biomarker values to lie within the [0,1] range. For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.

For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from our own controls and PCA subjects. As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of our own controls to be equal to the FA mean and standard deviation of the ADNI controls. No DTI data from PCA subjects was exposed to the algorithm at training time.

@@

\begin{figure}[htp]
\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
 \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories. (top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases. We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1"). For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunction scores $f(s_i;\lambda_d^l)$. Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunction scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
 \label{fig:dktSynthTrajCompTrue}
\end{figure}

%\begin{figure}
%\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
% \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories. Estimation of the trajectories in biomarkers 0,1,4 and 5 has been done without any data from the "synthetic PCA" disease, only based on the disease-agnostic correlations with biomarkers 2 and 3.}
% \label{fig:dktSynthTrajDrcSpace}
%\end{figure}

@@

% Fig1: biomarker traj. over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Results on TADPOLE dataset and our own dataset are presented in Fig. \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data. The model shows an unbiased data fit (Fig. \ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C). While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.

@@

\begin{picture}(5,5)
\put(0,20){\textbf{\Large{B}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% typical AD\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
% \caption{C}
\end{subfigure}
\begin{picture}(5,5)
\put(0,20){\textbf{\Large{C}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% PCA\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png} 
% \caption{PCA}
\end{subfigure}
\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}. Subject data from ADNI and our local cohort are also shown. The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject. The Y-axis measures biomarker values (normalised). (B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA, emulating top subfigures from Fig \ref{fig:diagram}.}
\label{fig:pcaTadDisSpace}
\end{figure}

@@

We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from our own cohort. We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers. Table \ref{sec:dktPerfMetrics} shows the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values. We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on. We also show the same performance metrics for two simpler models: (1) a \emph{latent stage model}, as described in \cite{jedynak2012computational}, which assumes all tAD and PCA subjects follow the same progression and (2) a \emph{linear univariate model} that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.

@@

We provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score). However, DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously. The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.

@@

Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our own centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.

There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions. On these subgroups, DKT can be used to transfer biomarker modalities left out during training.

As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g. AD variants), as well as across diseases (e.g. AD vs FTD). For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are. This can be done by choosing different a-priori allocations of biomarkers to agnostic units -- e.g. allocating amyloid and tau into one a shared agnostic unit, or otherwise directly to the disease specific models, thus making them dependent on the disease. The model with the best data fit or predictive power will indicate the most likely mechanism. For distinct neurodegenerative diseases, it is also currently unclear where their corresponding event-cascades merge, and the differences in pathological events across brain regions. DKT can be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units.

>>>>

\setlength{\textfloatsep}{15pt}
%\setlength{\floatsep}{1pt}
%\setlength{\topmargin}{1pt}

@@

\author{R\u{a}zvan V. Marinescu\inst{1} \and Marco Lorenzi\inst{1,4} \and Stefano Blumberg\inst{1} \and 
 Alexandra L. Young\inst{1} \and Pere P. Morell\inst{1} \and Neil P. Oxtoby\inst{1} \and Arman Eshaghi\inst{1,2} \and Keir X. Yong\inst{3} \and Sebastian J. Crutch\inst{3} \and Daniel C. Alexander\inst{1}, for 
the  Alzheimer’s  Disease  Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}

\authorrunning{R\u{a}zvan V. Marinescu et al.} % abbreviated author list (for running head)

\institute{Centre for Medical Image Computing, University College London, UK
\and 
Queen Square MS Centre, UCL Institute of Neurology, UK
\and 
Dementia Research Centre, University College London, UK
\and
University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
% F-91405 Orsay Cedex, France}
}

@@

%\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
\newcommand{\expFld}{.}

\begin{abstract}
We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases. DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases. DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases. As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks. Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects. For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC) UK, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available. We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature. We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches. While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases. Source code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.

@@

We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes. This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology. DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects at risk of diseases. We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data. This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK. We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings. We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches. Code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
 \caption{Diagram of the proposed DKT framework. We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions. Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region. For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe. The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled. Disease knowledge transfer can then be achieved via the disease-agnostic units. Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
 \label{fig:diagram}
\end{figure}

@@

\begin{equation}
\label{eqDysfunctionScoreDef}
 \gamma_i^l = f(\beta_i + m_{ij}; \lambda_{d_i}^l)
\end{equation}
where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of all agnostic units. Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled. For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.  We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers. Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.

@@

We first test DKT on synthetic data, to assess its performance when ground truth is known. More precisely, we generate data that follows the DKT model exactly, and test DKT's ability to recover biomarker trajectories and subject time-shifts. We generate synthetic data from two diseases (i.e. \emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE and DRC cohorts. The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$. To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects. Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g. MRI volume) but to represent measurements from different brain regions (e.g. temporal and occipital).

 \begin{table}
 %\fontsize{7}{12}
 \centering
 \begin{tabular}{c | c}
& \textbf{Trajectory parameters} \\
%\hline
 Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\

"Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
 "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
\hline
& \textbf{Subject parameters} \\
 Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
 Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
 Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
 Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
\end{tabular}
\caption{Parameters used for synthetic data generation, emulating the TADPOLE and DRC datasets.}
\label{tab:synParams}
\end{table}

@@

We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table. From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan. Most subjects also had MRI scans and cognitive tests. In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.

For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe. For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs DRC). Finally, we normalised the biomarker values to lie within the [0,1] range. For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.

For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from controls and PCA subjects from the DRC dataset. As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of the DRC controls to be equal to the FA mean and standard deviation of the ADNI controls. No DTI data from PCA subjects was exposed to the algorithm at training time.

@@

\begin{figure}[htp]
%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
\includegraphics[width=\textwidth]{compTrueParams101_synth1_JMD.pdf}
 \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories. (top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases. We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1"). For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunction scores $f(s_i;\lambda_d^l)$. Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunction scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
 \label{fig:dktSynthTrajCompTrue}
\end{figure}

%\begin{figure}
%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
% \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories. Estimation of the trajectories in biomarkers 0,1,4 and 5 has been done without any data from the "synthetic PCA" disease, only based on the disease-agnostic correlations with biomarkers 2 and 3.}
% \label{fig:dktSynthTrajDrcSpace}
%\end{figure}

@@

% Fig1: biomarker traj. over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Results on TADPOLE and DRC datasets are presented in Fig. \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data. The model shows an unbiased data fit (Fig. \ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C). While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.

@@

\begin{picture}(5,5)
\put(0,20){\textbf{\Large{B}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% typical AD\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
% \caption{C}
\end{subfigure}
\begin{picture}(5,5)
\put(0,20){\textbf{\Large{C}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% PCA\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png} 
% \caption{PCA}
\end{subfigure}
\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}. Subject data from ADNI and DRC cohorts are also shown. The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject. The Y-axis measures biomarker values (normalised). (B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA, emulating top subfigures from Fig \ref{fig:diagram}.}
\label{fig:pcaTadDisSpace}
\end{figure}

@@

We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from the DRC. We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers. Table \ref{sec:dktPerfMetrics} shows the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values. We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on. We also show the same performance metrics for two simpler models: (1) a \emph{latent stage model}, as described in \cite{jedynak2012computational}, which assumes all tAD and PCA subjects follow the same progression and (2) a \emph{linear univariate model} that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.

@@

While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously. The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.

@@

Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and the DRC centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.

There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.

As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g. AD variants), as well as across diseases (e.g. AD vs FTD). For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are. This can be done by choosing different a-priori allocations of biomarkers to agnostic units -- e.g. allocating amyloid and tau into one a shared agnostic unit, or otherwise directly to the disease specific models, thus making them dependent on the disease. The model with the best data fit or predictive power will indicate the most likely mechanism.
